TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-17 manuscript/manuscript.Rmd
Previously uncommitted changes

**Background.** Complex phenotypes such as height and intelligence, are thought to be a product of the collective effects of multiple phenotype-associated genes, or, more precisely, interactions among their protein products. High/low degree of interactions is suggestive of coherent/random molecular mechanisms, respectively. Comparing the degree of interactions may help to better predict and understand the coherence of the molecular mechanisms underlying complex phenotypes. However, direct comparison of the degree of interactions is difficult due to differences in size and configuration of phenotype-associated gene networks.

@@

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (with exemplary high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric diseases and intelligence-related traits were the least coherent.

@@

This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the ratio of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absense of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.

This study presents a novel measure of network coherence as a ratio of the normalized internal vs. external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used KEGG pathways [@Kanehisa:2000aa] as a reference for a paradigmatically high-coherent network [@Lee:2003aa], while sets of randomly sampled genes were used as the reference for the absence of coherence. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have the most coherence. Our method enables direct comparison of coherence measures and highlight gaps in current understanding of the molecular mechanisms of many phenotypes, e.g., schizophrenia being the lowest coherence disorder in the already low-coherence "Psychiatric disease" category.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by the interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensured that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs are removed from the analysis (see Methods). Second, unfiltered STRING data were used to maximize the use of PPI information at the expense of potential noise (referred hereafter as “STRING full”). The advantage of using unfiltered data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, were used. The use of different data sources (STRING and BioGrid) and filtering (full or filtered STRING data) was intended to increase the generalizability of our conclusions.

<!--Tables/table_1.csv-->
**Table 1. Summary statistics of the analyzed categories.** Average coherence estimates are shown. "NA" values indicate that a phenotype was not considered due to the low number of genes with PPIs.

@@

A typical network consists of nodes connected by edges [@Barabasi:2004aa]. In genomics, nodes typically represent genes and edges correspond to some measure of interactions, e.g., gene co-expression or interactions between protein products [@Goncalves:2012aa; @Wang:2012aa]. In terms of phenotypes, an intuitive expectation is that they will be represented by coherent modules of functionally related genes, similar gene expression profiles, shared genomic variants, higher PPI interactions, higher co-morbidity [@Zhang:2010ab; @Goh:2007aa; @Driel:2006aa; @Wang:2012aa; @Hamaneh:2015aa; @Menche:2015aa; @Jia:2011aa], reviewed in [@Dozmorov:2018aa; @Wang:2011aa]. In terms of graph theory, a coherent network is a community consisting of a group of nodes that are relatively highly connected to each other but sparsely connected to other nodes in the network [@fortunato2010community; @Radicchi:2004aa].

Numerous network properties have been defined to describe network structures [@Barabasi:2011aa; @Wang:2012aa]. A degree of a node, or connectivity, is one of the most fundamental characteristics defining the number of other nodes connected to a given node. A collection of degrees of all network nodes form a degree distribution. Comparing degree distributions among networks is an intuitive way to gain an understanding of network similarities and differences [@Barabasi:2004aa; @Feldman:2008aa]. We hypothesized that comparing degree distributions of networks formed by phenotype-associated gene sets may inform us about similarities and differences among phenotypes in terms of coherence of their antecedent and underlying molecular mechanisms.

@@

The coherence of phenotype-associated gene sets should be measured with respect to the reference level of networks with paradigmatically high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Supplementary Table S2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks of high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathwaays (REACTOME genes) are expected to interact more frequently. In our study, KEGG networks were used as the primary reference of high coherence, in parallel with GOCC and REACTOME collections. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of KEGG and random networks represent reference levels for high and low coherence, respectively.

@@

Direct comparison of degree distributions of the networks formed by phenotype-associated genes is hindered by the fact they depend on network size and configuration. An intuitive example is to consider a ring and a fully connected network of size 10 and 20 nodes, respectively (Figure 1). The degree distributions of the ring networks can be directly compared regardless of network size (vectors of "2" of length 10 and 20, respectively, Figure 1A).  Although the fully connected networks are similar in that they are fully connected (i.e., similarly coherent), their degree distributions differ (all nodes in the 10-node network have degree 9, while all nodes in the 20-node network have degree 19, Figure 1B). <!--Normalizing the degree distributions of the fully connected networks by dividing them by network size makes them identical as would be expected for networks with similar coherence. However, similar normalization of ring networks will make their degree distributions different (0.22 for the 10-node network and 0.11 for the 20-node network), leading to the illusory impression of differences in coherence between intuitively equally coherent networks. -->These observations highlight the difficulty in comparing degree distributions of phenotype-associated gene networks due to two facts: 1) network sizes differ, and 2) network configuration (fully connected, ring, or intermediate connectivity) is unknown.

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed negative correlation of network size with the slopes of KEGG and random networks (-0.37/-0.85, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the ten KEGG pathways closest in size and ten random networks matched in size as references for high/low coherence for the corresponding phenotype. Consequently, we use the slopes of internal vs. external degree distributions created by these size-matched references to normalize the slope of the phenotype-specific network to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed removes the dependency between coherence and network size (Supplementary Figure S1).

@@

To estimate the significance of coherence measures, we developed a permutation-based approach (see Methods). Notably, we observed coherence for all diseases to be highly significant (permutation p-value < 0.002, Supplementary Table S3). The distribution of coherence estimates was skewed towards 1.00 (highest coherence), with the median coherence being 0.86. "Obesity-related traits" and "Height" were on the lowest range of coherence estimates (0.42 and 0.54, respectively). "Weight", "Telomere length", and "Corneal structure" were on the highest range of coherence (1.00, Supplementary Table S3). These results indicate that coherence of phenotype-specific networks is generally high and significantly different from random.

@@

We estimated coherence of 53 traits from five categories (Figure 3, Supplementary Figures S3-S5). Among anthropometric traits, "Weight" and "Body Mass Index" had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose- and cholesterol-related traits (metabolic). Notably, metabolic and eye trait categories had the highest overall coherence (mean coherence 0.90 and 0.93, respectively). Several notable examples of low coherence include "Obesity-related traits", "Height", and "Longevity" (anthropometric traits), blood count-related traits (cardiovascular). Traits in the "Intelligence" category had low overall coherence (mean coherence = 0.76). These results highlight the relatively good knowledge of the genetics of metabolic traits which corresponds to highly coherent networks. Conversely, current knowledge about the genetics of intelligence-related traits appear insufficient as reflected in the low coherence of the corresponding networks.

@@

Human protein-protein interactions (PPIs) were obtained from STRING database v.11.0 [@Szklarczyk:2015aa] and BioGrid database v.3.5.174 [@Chatr-Aryamontri:2017aa] (accessed 06/25/2019). The STRING data for the tax ID "9606" (human) was used, and Ensembl protein IDs were converted to gene names using BioMart. Either the full dataset (11,761,040  entries) or data filtered by "combined_score" > 500 (range 0-1000) (1,373,946 entries, the main PPI dataset) was used. The BioGrid data was filtered to include "TaxID interactor" type equal to "9606", all "Interaction Types" were included, totaling 465,660 interactions. These three PPI databases were used in parallel to evaluate disease coherence.

@@

**Supplementary Table S1. Summary statistics of the analyzed phenotypes.** Each worksheet corresponds to the PPI database used.

**Supplementary Table S2. Summary of networks from MSigDB categories.** Minimum, maximum, mean and median of the slopes and number of genes are shown for each collection, along with the total number of networks and the correlation between network size and slope.

**Supplementary Table S3. Phenotype-specific coherence estimates.** The PPI database (STRING, STRING filtered, Biogrid) and the reference of high coherence (KEGG, GOCC, Reactome) are specified in the corresponding columns for coherence and permutation p-value estimations.

**Supplementary Figure S1. Normalized coherence is unaffected by network size.** Red/blue gradient and numbers represent Pearson correlation coefficients between corresponding entries.

**Supplementary Figure S2. Coherence estimates are similar when using different references of high coherences and PPI databases.** Red/blue gradient and numbers represent Pearson correlation coefficients between corresponding entries.

**Supplementary Figure S3. Coherence estimates of traits.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis). Missing entries indicate that for a given PPI database a phenotype-associated network could not be built.

>>>>

**Background.** Complex phenotypes such as height and intelligence, are thought to be a product of the collective effects of multiple phenotype-associated genes, or, more precisely, interactions among their protein products. High/low degree of interactions is suggestive of coherent/random molecular mechanisms, respectively. Comparing the degree of interactions may help to better understand the coherence of the molecular mechanisms underlying complex phenotypes and the potential for therapeutic intervention. However, direct comparison of the degree of interactions is difficult due to differences in size and configuration of phenotype-associated gene networks.

@@

**Results.** Using STRING and BioGrid PPI databases, we compared the coherence of 133 phenotype-associated gene sets from GWAScatalog against size-matched KEGG pathways (the reference for high coherence) and random networks (the lower limit of coherence). We observed a range of coherence estimates for each category of phenotypes. Metabolic traits and diseases were the most coherent, while psychiatric diseases and intelligence-related traits were the least coherent.

@@

<!--This study compares the connectivity of networks formed by disease- or trait-associated genes and gene products. We hypothesized that genes associated with (categories of) diseases share common molecular mechanisms and, hence, form connected (i.e., coherent) networks. Therefore, the level of connectivity may inform us about similarities and differences among the diseases. The terms “network connectivity,” “coherence,” or “average degree distribution” are used interchangeably as appropriate. We developed a method for normalizing the internal network connectivity that accounts for differences in network sizes formed by disease-associated genes. We compared the ratio of normalized internal vs. external network connectivity between psychiatric, neurologic, autoimmune, cancer, cardiovascular, metabolic and other diseases and traits. We used KEGG pathways [30] as a reference for a coherent network with common molecular mechanisms, and we used sets of randomly sampled genes as a reference for the absense of coherence. We found that genes associated with metabolic diseases and traits and autoimmune diseases formed coherent networks similar to KEGG pathways. On the contrary, the coherence of networks of psychiatric and neurologic disease-associated genes, as well as intelligence-associated traits, was similar to that of randomly selected genes. Cancer, anthropometric traits, eye-, and cardiovascular diseases and traits formed networks of intermediate connectivity. Our study identifies significant differences between categories of diseases in terms of network properties of the disease-associated genes and highlights the limitations of the current understanding of the molecular determinants of psychiatric disorders.-->
This study presents a novel measure of network coherence as a ratio of the internal vs. external degree distributions. The internal degree distribution is defined as gene-specific interaction counts within a protein-protein interaction (PPI) network formed by phenotype-associated genes, while the external degree distribution counts interactions with other genes in the whole PPI network. We used KEGG pathways [@Kanehisa:2000aa] as a reference for a paradigmatically high-coherent network [@Lee:2003aa], while sets of randomly sampled genes were used as the reference for the absence of coherence. Using KEGG and random networks matched in size to a phenotype-specific network, we derived normalized measure of coherence that can be compared across phenotypes. We hypothesized that the level of coherence may inform us about similarities and differences among phenotypes. Using two PPI databases (STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa]), we compared coherence of 133 phenotypes from GWAScatalog [@Hindorff:2009aa; @MacArthur:2017aa]. Our results show the tendency of complex phenotypes, such as intelligence-related traits, to have low coherence and metabolic traits and diseases to have high coherence. Our method enables direct comparison of coherence measures and highlight gaps in current understanding of the molecular mechanisms of many phenotypes, e.g., schizophrenia being the lowest coherence disorder in the already low-coherence "Psychiatric disease" category.

@@

We created networks of phenotype-associated genes using PPI information from STRING [@Szklarczyk:2015aa] and BioGrid [@Chatr-Aryamontri:2017aa] databases (see Methods). We opted for the use of three types of PPI data. First, STRING data filtered by the interaction confidence score above 500 (middle of the bimodal distribution of the score, 0-1000 range, referred hereafter as “STRING filtered”) was used as the primary source of curated PPI data. This filtering step ensured that only high-confidence PPIs were selected. The advantage of using filtered data is the reliance on high-quality PPIs. The disadvantage is that the number of PPIs may be insufficient for forming phenotype-associated networks; consequently, phenotypes with genes without PPIs are removed from the analysis (see Methods). Second, complete STRING data were used to maximize the use of PPI information at the expense of potential noise (referred hereafter as “STRING”). The advantage of using complete data is that more trait-associated genes will have PPI information and, hence, can be analyzed. The disadvantage, however, is that the results may be less reliable due to the presence of noisy PPIs having low confidence scores. Third, BioGrid PPI data, which has been curated to keep only high-confidence interactions, were used. The use of different data sources (STRING and BioGrid) and filtering (complete or filtered STRING data) was intended to increase the generalizability of our conclusions.

<!--Tables/table_1.csv-->
**Table 1. Summary statistics of the analyzed categories.** Average coherence estimates are shown. "NA" values indicate that a category lacked phenotypes with sufficient number of PPIs. Sorted by "String Filtered" average coherence.

@@

A typical network consists of nodes connected by edges [@Barabasi:2004aa]. In genomics, nodes typically represent genes and edges correspond to some measure of interactions, e.g., gene co-expression or interactions between protein products [@Goncalves:2012aa; @Wang:2012aa]. In terms of phenotypes, an intuitive expectation is that they will be represented by coherent networks of functionally related genes, similar gene expression profiles, shared genomic variants, higher PPI interactions, higher co-morbidity [@Zhang:2010ab; @Goh:2007aa; @Driel:2006aa; @Wang:2012aa; @Hamaneh:2015aa; @Menche:2015aa; @Jia:2011aa], reviewed in [@Dozmorov:2018aa; @Wang:2011aa]. In terms of graph theory, a coherent network is a community consisting of a group of nodes that are relatively highly connected to each other but sparsely connected to other nodes in the network [@fortunato2010community; @Radicchi:2004aa].

Numerous network properties have been defined to describe network structures [@Barabasi:2011aa; @Wang:2012aa]. A degree of a node, or connectivity, is one of the most fundamental characteristics defining the number of other nodes connected to a given node. A collection of degrees of all network nodes form a degree distribution. Comparing degree distributions among networks is an intuitive way to gain an understanding of network similarities and differences in terms of connectivity (aka coherence) of the corresponding genes [@Barabasi:2004aa; @Feldman:2008aa]. We hypothesized that comparing degree distributions of networks formed by phenotype-associated gene sets may inform us about similarities and differences among phenotypes in terms of coherence estimates, informing us about the underlying molecular mechanisms.

@@

The coherence of phenotype-associated gene sets should be measured with respect to the reference level of networks with paradigmatically high and low coherence. To establish a reference of high coherence, we considered networks from collections of the Molecular Signatures Database (MSigDB v6.2) [@Subramanian:2005aa]. MSigDB contains sets of genes having various types of functional relationships. We found that pathways from the Kyoto Encyclopedia of Genes and Genomes (KEGG) [@Kanehisa:2000aa] had the smallest average slope corresponding to the highest level of coherence (Supplementary Table S2). Networks assembled from Gene Ontology - Cellular Component (GOCC) collection and REACTOME pathways followed, representing alternative sources of networks of high coherence. This is expected as genes expressed in the same cellular components (GOCC genes) or participating in the same metabolic pathwaays (REACTOME genes) are presumed to interact more frequently. In our study, KEGG networks were used as the primary reference of high coherence, in parallel with GOCC and REACTOME collections. Conversely, as a reference for low coherence, we randomly sampled genes from a pool of genes in a given PPI database. These random gene lists represent expected network coherence arising by chance. The slopes of KEGG and random networks represent reference levels for high and low coherence, respectively.

@@

Direct comparison of degree distributions of the networks formed by phenotype-associated genes is hindered by the fact they depend on network size and configuration. An intuitive example is to consider a ring and a fully connected network of size 10 and 20 nodes, respectively (Figure 1). The degree distributions of the ring networks can be directly compared regardless of network size (vectors of "2" of length 10 and 20, respectively, Figure 1A).  Although the fully connected networks are similar in that they are fully connected (i.e., similarly coherent), their degree distributions differ (all nodes in the 10-node network have degree "9", while all nodes in the 20-node network have degree "19", Figure 1B). <!--Normalizing the degree distributions of the fully connected networks by dividing them by network size makes them identical as would be expected for networks with similar coherence. However, similar normalization of ring networks will make their degree distributions different (0.22 for the 10-node network and 0.11 for the 20-node network), leading to the illusory impression of differences in coherence between intuitively equally coherent networks. -->These observations highlight the difficulty in comparing degree distributions of phenotype-associated gene networks due to two facts: 1) network sizes differ, and 2) network configuration (fully connected, ring, or intermediate connectivity) is unknown.

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed negative correlation of network size with the slopes of KEGG and random networks (-0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the ten KEGG pathways closest in size and ten random networks matched in size as references for high/low coherence for the corresponding phenotype. Consequently, we use the slopes of internal vs. external degree distributions created by these size-matched references to normalize the slope of the phenotype-specific network to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed removes the dependency between coherence and network size (Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

To estimate the significance of coherence measures, we developed a permutation-based approach (see Methods). Notably, we observed coherence for all diseases to be highly significantly different from random (permutation p-value < 0.002, Supplementary Table S3). Similarly, the distribution of coherence estimates was skewed towards 1.00 (highest coherence), with the median coherence being 0.86. Among traits, "Corneal structure", "Body mass index", "Weight" were the most coherent (1.02, 1.00, 0.99, respectively). [???]

"Obesity-related traits" and "Height" were on the lowest range of coherence estimates (0.42 and 0.54, respectively). "Weight", "Telomere length", and "Corneal structure" were on the highest range of coherence (1.00, Supplementary Table S3). These results indicate that coherence of phenotype-specific networks is generally high and significantly different from random.

@@

We estimated coherence of 54 traits from five categories (Figure 3, Supplementary Table S3, Supplementary Figures S3-S5). Among anthropometric traits, "Weight" and "Body Mass Index" had the highest level of coherence. Other examples of traits with high coherence include blood pressure-related traits (cardiovascular) and glucose- and cholesterol-related traits (metabolic). Notably, metabolic and eye trait categories had the highest overall coherence (mean coherence 0.90 and 0.93, respectively). Several notable examples of low coherence include "Obesity-related traits", "Height", and "Longevity" (anthropometric traits), blood count-related traits (cardiovascular). Traits in the "Intelligence" category had low overall coherence (mean coherence = 0.76). These results highlight the relatively good knowledge of the genetics of metabolic traits which corresponds to highly coherent networks. Conversely, current knowledge about the genetics of intelligence-related traits appear insufficient as reflected in the low coherence of the corresponding networks.

@@

Human protein-protein interactions (PPIs) were obtained from STRING database v.11.0 [@Szklarczyk:2015aa] and BioGrid database v.3.5.174 [@Chatr-Aryamontri:2017aa] (accessed 06/25/2019). The STRING data for the tax ID "9606" (human) was used, and Ensembl protein IDs were converted to gene names using BioMart. Either the complete dataset (11,761,040  entries) or data filtered by "combined_score" > 500 (range 0-1000) (1,373,946 entries, the main PPI dataset) was used. The BioGrid data was filtered to include "TaxID interactor" type equal to "9606", all "Interaction Types" were included, totaling 465,660 interactions. These three PPI databases were used in parallel to evaluate disease coherence.

@@

**Supplementary Table S1. Summary statistics of the analyzed phenotypes.** Minimum, mean, median, maximum of coherence estimates, the number of genes, SNPs, the total number of phenotype networks. Each worksheet corresponds to the PPI database used.

**Supplementary Table S2. Summary of networks from MSigDB categories.** Minimum, mean, median, maximum of the slopes, the number of genes, and the total number of networks are shown for each collection, along with the correlation between network size and slope. "CP:/KEGG/Reactome/Biocarta" - canonical pathways, "GOBP/GOMF/GOCC" - gene ontology biological processes, molecular functions, cellular component, "TFT" - transcription factor targets, "CGN" - cancer gene neighborhoods, "CM" - cancer modules, "MIR" - microRNA targets, "CGP" - chemical and genetic perturbations.

**Supplementary Table S3. Phenotype-specific coherence estimates.** The PPI database (STRING, STRING filtered, Biogrid) and the reference of high coherence (KEGG, GOCC, Reactome) are specified in the corresponding columns for coherence. Permutation p-values were estimated using STRING filtered database and KEGG reference of high coherence. "Trait" and "Disease" sheets correspond to subsets of the main table containing coherence estimates for traits/diseases, respectively, sorted by descending "String Filt Coherence KEGG" column.

**Supplementary Figure S1. Normalized coherence is unaffected by network size.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, and the normalized coherence estimates using the corresponding PPI databases and KEGG pathways as a reference of high coherence.

**Supplementary Figure S2. Coherence estimates are similar when using different references of high coherences and PPI databases.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the normalized coherence estimates using the corresponding PPI databases and KEGG/Reactome pathways and Gene Ontology Cellular Component networks as a reference of high coherence.

**Supplementary Figure S3. Coherence estimates of traits using KEGG as a reference of high coherence.** Size of dots represents the level of normalized coherence (X-axis) for individual traits (Y-axis). Plots are faceted by PPI database used to estimate coherence. Missing entries indicate that, for a given PPI database, a phenotype-associated network could not be built.

**Supplementary Figure S4. Coherence estimates of traits using GOCC as a reference of high coherence.** See legend for Supplementary Figure S3.

**Supplementary Figure S5. Coherence estimates of traits using Reactome as a reference of high coherence.** See legend for Supplementary Figure S3.

**Supplementary Figure S6. Coherence estimates of diseases using KEGG as a reference of high coherence.** See legend for Supplementary Figure S3.

**Supplementary Figure S7. Coherence estimates of diseases using GOCC as a reference of high coherence.** See legend for Supplementary Figure S3.

**Supplementary Figure S8. Coherence estimates of diseases using Reactome as a reference of high coherence.** See legend for Supplementary Figure S3.
